---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r loading, echo = TRUE}
  library(tidyverse)
  activity <- read_csv("activity.csv")
  head(activity)
```

## What is mean total number of steps taken per day?

Calculate total number of steps per day

```{r steps_per_day, echo = TRUE}
steps_per_day <- with(activity, aggregate(steps, list(date), FUN = sum, na.rm = TRUE))
head(steps_per_day)
```

Show a histogram

```{r histogram_steps_per_day, echo = TRUE}
hist(x = steps_per_day$x, main = "Histogram of Steps per Day", xlab = "Total number of Steps", ylab = "Frequency")

abline(v = mean(steps_per_day$x),                       # Add line for mean
       col = "red",
       lwd = 3)
text(x = mean(steps_per_day$x) / 2.2,                   # Add text for mean
     y = 20 ,
     paste("Mean = \n", mean(steps_per_day$x)),
     col = "red",
     cex = 1.5)
```

### Mean and median number of steps taken each day

Mean (excluding missing values)

```{r calc-mean-steps-per-day, echo = TRUE}
mean(steps_per_day$x, na.rm = TRUE)
```

Median (excluding missing values)

```{r calc-median-steps-per-day, echo = TRUE}
median(steps_per_day$x, na.rm = TRUE)
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo = TRUE}
activity%>% group_by(interval) %>%
    summarize(meanByInterval = mean(steps, na.rm = TRUE)) %>%
    filter(meanByInterval == max(meanByInterval))
```

## What is the average daily activity pattern?

```{r average_daily_pattern, echo = TRUE}
  stepsPerIntervalAvg <- aggregate(steps ~ interval, data = activity, FUN = mean, na.action = na.omit)
colnames(stepsPerIntervalAvg) <- c("interval", "steps")
ggplot(stepsPerIntervalAvg) + 
  geom_line(aes(interval,steps)) +
  ggtitle("Average daily Pattern - Steps per Interval")
```



One can clearly see that the Person sleeps in the morning, does the most activity on 09:00 AM, than has mixed activity until 19:00 PM and then the activity decreases steadily.

## Imputing missing values

Let's see, which days are affected by "NA"s

```{r steps_per_day_with_NA, echo = TRUE}
steps_per_day <- with(activity, aggregate(steps, list(date), FUN = sum, na.rm = FALSE))
head(steps_per_day)
```


```{r plot_na, echo = TRUE}
par(mfrow=c(2,1))    # set the par-function for subplots
plot(steps_per_day$x ~ steps_per_day$Group.1, xlab = "steps per day")
barplot(is.na(steps_per_day$x), col = "red", ylab = "has NA")

```

So how many NA have we got?

```{r, echo = TRUE}
sum(is.na(activity$steps) == TRUE)
```
okay, enough to take care about.

### actual imputation
strategy: fill the void with the mean value in that 5-Minute-Timeframe

```{r imputation, echo = TRUE}
activity_imp <- activity # create new dataset
for (i in stepsPerIntervalAvg$interval) {
    activity_imp[activity_imp$interval == i & is.na(activity_imp$steps), ]$steps <- 
        stepsPerIntervalAvg$steps[stepsPerIntervalAvg$interval == i]
}
head(activity_imp) # no NAs
```
### plot imputed data

````{r plot imputed data, echo = TRUE}
steps_per_day_imp <- with(activity_imp, aggregate(steps, list(date), FUN = sum, na.rm = TRUE))
hist(x = steps_per_day_imp$x, main = "Histogram of Steps per Day Imputed", xlab = "Total number of Steps", ylab = "Frequency")

abline(v = mean(steps_per_day_imp$x),                       # Add line for mean
       col = "red",
       lwd = 3)
text(x = mean(steps_per_day_imp$x) / 2.2,                   # Add text for mean
     y = 20 ,
     paste("Mean = \n", mean(steps_per_day_imp$x)),
     col = "red",
     cex = 1.5)

```
Mean

```{r, echo = TRUE}
mean(steps_per_day_imp$x)
```

Median

```{r, echo = TRUE}
median(steps_per_day_imp$x)
```
Finding: Yes, that imputation strategy has a slight effect on the outcome.

## Are there differences in activity patterns between weekdays and weekends?

firstly, we have to determine, which date is weekend and weekday.
We can use the function `format(as.Date(df$date),"%w")` to do so. It return numbers from 0-6 for Monday to Sunday.

```{r, determine_weekdays, echo = TRUE}
activity %>% add_column(weekend = as.integer(format(as.Date(activity$date),"%w"))>=4) -> activity_weekdays
 
stepsPerIntervalAvg_weekdays <- aggregate(steps ~ interval+ weekend, data = activity_weekdays, FUN = mean, na.action = na.omit)

ggplot(stepsPerIntervalAvg_weekdays) + 
  geom_line(aes(interval,steps, col = weekend)) +
  ggtitle("Average daily Pattern - Steps per Interval")
```


One can clearly see that the person get up on weekdays earlier. Another finding is that the person does a Walk at 15:00 PM on Weekend.

